---
title: "Phase1"
output: html_notebook
---

Installing the packages


```{r}
install.packages("scatterplot3d") 
install.packages("dplyr")
install.packages("ggplot2")
```

Reading libraries and the dataset

```{r}
library(ggplot2)
library(scatterplot3d)
library(dplyr)
library(readr)
dataset <- read_csv("Dataset/students_adaptability_level_online_education.csv")
```

Sample of raw dataset and summary 
```{r}
View(dataset)
summary(dataset)
```
 Missing values
```{r}
sum(is.na(dataset))
```
Since there are no missing values, and since there are no numerical data to be checked for outliers, we will move to the next step

Data summary
```{r}
summary(dataset$Age)
summary(dataset$Gender)
summary(dataset$Device)
summary(dataset$Location)
summary(dataset$'Self Lms')
summary(dataset$'IT Student')
summary(dataset$'Network Type')
summary(dataset$'Internet Type')
summary(dataset$'Load-shedding')
summary(dataset$'Class Duration')
summary(dataset$'Education Level')
summary(dataset$'Adaptivity Level')
summary(dataset$'Institution Type')
summary(dataset$'Financial Condition')
```
All attribute types are char, so we can't calculate variance or discretize.


Converting data types from char to factors so they can be plotted
```{r}
dataset$'Financial Condition' <- as.factor(dataset$'Financial Condition')
dataset$'Institution Type' <- as.factor(dataset$'Institution Type')
dataset$'Education Level' <- as.factor(dataset$'Education Level')
dataset$'Load-shedding' <- as.factor(dataset$'Load-shedding')
dataset$'Internet Type'<- as.factor(dataset$'Internet Type')
dataset$'Network Type'<- as.factor(dataset$'Network Type')
dataset$Device <- as.factor(dataset$Device)
dataset$Gender<- as.factor(dataset$Gender)
dataset$Age <- as.factor(dataset$Age)
```

Plotting Methods:

Bar plot of the age attribute, which is a range
```{r}
ggplot(dataset, aes(x = Age)) + geom_bar()
```

Mosaic plot of the educational level attribute with the financial condition
```{r}
counts <- table(dataset$'Education Level', dataset$'Financial Condition')
mosaicplot(counts, xlab='Education Level', ylab='Financial Condition',
           main='Education Level by Financial Condition', col='forestgreen')
```

Scatter plots of the gender and device and class duration attributes
and internet type and load-shedding and institution type attributes
```{r}
scatterplot3d(dataset$Gender,dataset$Device, dataset$`Class Duration`)
scatterplot3d(dataset$'Internet Type',dataset$'Load-shedding', dataset$'Institution Type')
```


Pie chart for the adaptivity level attribute
```{r}
tab <- dataset$'Adaptivity Level'%>% table()
precentages <- tab %>% prop.table() %>% round(3) * 100 
txt <- paste0(names(tab), '\n', precentages, '%') 
pie(tab, labels=txt) 
```

Since this dataset does not contain numerical values, discreization and normalization cannot be performed.